---
title: "Class 11: Comparative Structure Analysis"
author: "Allen (A16897142)"
format: html
toc: True
---

```{r}
library(bio3d)
```
We will start by analyzing ADK. It is an inhibitor in cancer cells and a great area of research currently.The accession code is "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

Now we search on the PDB database to find the related entries,
```{r}
blast <- blast.pdb(aa)
```
```{r}
attributes(blast)
```
```{r}
head(blast$hit.tbl)
```

Make a summary figure with these results abve:
```{r}
hits <- plot(blast)
```

The top hits are our most similar entries in the results.
```{r}
hits$pdb.id
```

Downloading the pdb information:
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
Side Note:
```{r}
library(bio3dview)

view.pdbs(pdbs)
```

This view is better than the data we recieved previously, however it is still hard to see the alignment. It is hard to see which family of sequences are similar.

Let's do a PCA:
```{r}
pc <- pca(pdbs)
plot(pc)
```
```{r}
plot(pc,pc.axes=1:2)
```

Here is a PDB trajectory for mol-star
```{r}
mktrj(pc,file="pca_results.pdb")
```

